Neon: Nuclear Norm to Beat Muon

### 2025-10-13 poster/old/neon_poster.tex

\documentclass[a0]{a0poster}
% You might find the 'draft' option to a0 poster useful if you have
% lots of graphics, because they can take some time to process and
% display. (\documentclass[a0,draft]{a0poster})
\input defs
\pagestyle{empty}
\setcounter{secnumdepth}{0}
\renewcommand{\familydefault}{\sfdefault}
\newcommand{\QED}{~~\rule[-1pt]{8pt}{8pt}}\def\qed{\QED}

@@

\hrule\medskip
\Head{Introduction}\\
Recent advances in optimization techniques have highlighted the benefits of leveraging the matrix structure of neural network weights during training. Optimizers like Muon (Jordan et al.) and Shampoo (Gupta, Koren and Singer) have shown promise, but at a significantly higher computational cost than traditional methods like Adam. To bridge this gap, we propose Neon, a new optimizer that builds upon the framework of Bernstein and Newhouse. By using alternative norms, such as nuclear norm (Nuclear-Neon) or custom $F*$ norm ($F*$-Neon), we induce low-rank update matrices that enable more efficient computation. We evaluate the performance of Neon, Muon, and Adam on MLP, CIFAR10, and NanoGPT, and demonstrate [resting results here].

@@

% \begin{figure}[h]
%     \center{\includegraphics[width=0.8\linewidth]{../../figs/mlp24/loss_val.png}}
%     \caption{NanoGPT validation loss}
%     \label{fig:nanogpt_loss}
% \end{figure} 
% \begin{figure}[h]
%     \center{\includegraphics[width=0.8\linewidth]{../../figs/nanogpt/train_loss.pdf}}
%     \caption{NanoGPT train loss}
%     \label{fig:nanogpt_train_loss}
% \end{figure}
\begin{figure}[h]
    \center{\includegraphics[width=0.8\linewidth]{../figs/nanogpt/val_loss.pdf}}
    \caption{NanoGPT validation loss}
    \label{fig:nanogpt_val_loss}
\end{figure}
\medskip
\hrule\medskip
\Head{Conclusion}\\
A new norm-motivated algorithm proved to be not so effective as Muon, at least from the first sight. However, its updates seem to be much faster to compute, and more so for a large matrices which has not tested yet. We believe that our algorithm is not an alternative to Muon, but probably a more effective tool for solving some other optimization problem. A deep inspection of Nuclear Norm and its properties might give us an insight into Neon's future.

>>>>

\documentclass[a1]{a1poster}
% You might find the 'draft' option to a0 poster useful if you have
% lots of graphics, because they can take some time to process and
% display. (\documentclass[a0,draft]{a0poster})
\input defs
\pagestyle{empty}
\setcounter{secnumdepth}{0}
\renewcommand{\familydefault}{\sfdefault}
\newcommand{\QED}{~~\rule[-1pt]{8pt}{8pt}}\def\qed{\QED}

@@

\hrule\medskip
\Head{Introduction}\\
Recent advances in optimization techniques have highlighted the benefits of leveraging the matrix structure of neural network weights during training. Optimizers like Muon (Jordan et al.) and Shampoo (Gupta, Koren and Singer) have shown promise, but at a significantly higher computational cost than traditional methods like Adam. To bridge this gap, we propose Neon, a new optimizer that builds upon the framework of Bernstein and Newhouse. By using alternative norms, such as nuclear norm (Nuclear-Neon) or custom $F*$ norm ($F*$-Neon), we obtain low-rank update matrices that enable more efficient computation. We evaluate the performance of Neon, Muon, and Adam on training MLP and CNN on CIFAR-10 and on finetuning NanoGPT.

@@

% \begin{figure}[h]
%     \center{\includegraphics[width=0.8\linewidth]{../../figs/mlp24/loss_val.png}}
%     \caption{NanoGPT validation loss}
%     \label{fig:nanogpt_loss}
% \end{figure} 
% \begin{figure}[h]
%     \center{\includegraphics[width=0.8\linewidth]{../../figs/nanogpt/train_loss.pdf}}
%     \caption{NanoGPT train loss}
%     \label{fig:nanogpt_train_loss}
% \end{figure}
\begin{figure}[h]
    \center{\includegraphics[width=0.8\linewidth]{../figs/nanogpt/val_loss.pdf}}
    \caption{NanoGPT validation loss}
    \label{fig:nanogpt_val_loss}
\end{figure}
\medskip
\hrule\medskip
\Head{Conclusion}\\
A new norm-motivated algorithm proved to be not so effective as Muon, at least from the first sight. However, its updates seem to be much faster to compute, and more so for large matrices that have not been tested yet. We believe our algorithm is not an alternative to Muon, but probably a more effective tool for solving some other optimization problems. A deep inspection of nuclear norm and its properties might give us an insight into Neon's future.
